I have received a great deal of support and assistance throughout the writing of this dissertation. I first want to thank my advisor, Prof. Ella Aktins, for her unwavering support these last five years. You have consistently helped me every week in research, writing, and even classwork! Whenever I doubted myself or struggled you helped me step back and see the bigger picture of my research goals and work. You paved the runway and provided the wind for my own research to take flight.

I would like to thank my committee members for valuable feedback on my work and ideas. Prof. Sarter, thank you for your encouragement to persevere in adversity and teaching me that rejection is par for the course. Prof. Stout, thank you for instilling in me a passion for writing efficient software and providing the knowledge to create parallel programs. Thank you Prof. Ghaffari for having an open ear and giving me opportunities in academic service.

% I would also like to thank the University of Michigan's many supportive services such as Student Legal Services, Dean of Students Office, and the Counseling and Psychological Services. Each of these programs had a hand in providing me incredible guidance and support.

Thank you to the past teachers and mentors who have supported me to this point. Prof. Hedengren, thank you for hiring me as an undergraduate assistant which began my journey to pursue higher education. Mr. Lenski (Jerry), thank you for being a wonderful mentor and teaching me everything I know about industrial control systems. You taught me that with patience and careful planning even seemingly insurmountable engineering tasks can be accomplished. 
% and giving me my first research and publication experience 

Thank you to everyone in the A2SYS lab (in the order I met them): Pedro, Mia, Hossein, Cosme, Brian, Prashin, Matt, Prince, Joseph, Paul, Mark, Anne, and John.  It has been a great honor in my life getting to know each of you. Matt, Prince, and Prashin, I will always remember the hackathons we joined and the thrill of competing together. Cosme and Brian, you are wonderful friends and research partners. Thank you for being a great example for me and listening to my crazy ideas.

Next I want to thank my family. Mom, you shine as an example of love and service. Thank you for all your support especially during the vulnerable parts of my childhood. To my siblings JD and Meghan, thank you for your love, affection, and friendship. To my kids Kiara, Emily, and Alan: you are treasures in my life and bring me joy every day. It is a privilege to be your Guardian and learn from you.  Finally, I want to acknowledge my wife, Natalia. We have taken this journey together. Whenever life became stressful and I faltered, you always stepped up and supported me through adversity. I love you completely. I am so excited for our next journey!  
%We have been tested. Yet we stand here, hand-in-hand, stronger than ever. I love you 


\paragraph{Funding}
This work was supported by the Rackham Merit Fellowship Program and NSF I/UCRC Award 1738714. 

